Linear vertical shear instability in astrophysical disks: a semi-numerical study

### 2014-12-17 linear_adia.tex
minor updates

\subsection{Adiabatic perturbations} 
We now extend the above discussion to general $\gamma$, but in the
adiabatic limit ($t_c\to\infty$). We eliminate variables from 
Eq. \ref{lin_mass}---\ref{lin_energy} in favor of $\delta v_z$ 
and $\Delta \equiv \nabla\cdot\delta\bm{v}$ to obtain
\begin{align}
  &\Delta\left(1 + \frac{\gamma c_s^2 k_x^2}{D}\right) = \frac{d\delta
    v_z}{dz} + \frac{\ii k_x^2}{D}\left(\ii c_s^2 \frac{d\ln\rho}{dz}
    - \frac{r}{k_x}\frac{d\Omega^2}{dz}\right)\delta v_z,\label{adia_iso1}\\
  & -\sigma^2\delta v_z = \gamma c_s^2 \frac{d\Delta}{dz} +
  c_s^2\frac{d}{dz}\left(\frac{d\ln\rho}{dz}\delta v_z\right) +
  \left(\gamma-1\right)\frac{d\ln\rho}{dz} \Delta.\label{adia_iso2}
\end{align}
{\bf note: incompressible mode, set delta to zero. only works for
  constant gravity.}

We again make the low-frequency approximation
and ignore the vertical dependence of $\Omega$ and $\kappa$ except
where their vertical derivatives appear explicitly in the above
equations. We thus make the replacement $D\to \Omega_k^2$, as before
for isothermal perturbations. Here, we also use the fact that in the
global disk, 
\begin{align}
  r\frac{\p \Omega^2}{\p z} = -\frac{\p \ln\rho}{\p z} \frac{qc_s^2}{r},
\end{align}
for the power-law prescription of the midplane sound-speed for
$\Gamma=1$.

Combining Eq. \ref{adia_iso1} and Eq. \ref{adia_iso2} with the above
substitutions we obtain, in terms of dimensionless variables defined
previously, an equation for $\delta v_z$, 
\begin{align}
  0 =& \dd v_z^{\prime\prime} + \left(1 + \ii \epsilon q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime\notag\\
  &+\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right)\right.\notag\\
  &\phantom{0=}\left.-\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii\epsilon  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}
We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  &\hat{\sigma}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat \notag\\
  &=  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
  &\phantom{=} +
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii\epsilon
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat\notag\\
  &\phantom{=} + \ii\epsilon q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
$\hat{\sigma}^2$ is real, and  $\hat{\sigma}^2>0$ if $\gamma>1$. As
expected, a sub-adiabatically stratified disk is stable in the absense
of vertical shear. In the presence of vertical shear $q\neq0$ and
$\hat{\sigma}$ is generally complex, which implies the possibility of
instability.

>>>>

\subsection{Adiabatic perturbations} 
We now extend the above discussion to general $\gamma$, but in the
adiabatic limit ($t_c\to\infty$). We eliminate variables from 
Eq. \ref{lin_mass}---\ref{lin_energy} in favor of $\delta v_z$ 
and $\Delta \equiv \nabla\cdot\delta\bm{v}$ to obtain
\begin{align}
  &\Delta\left(1 + \frac{\gamma c_s^2 k_x^2}{D}\right) = \frac{d\delta
    v_z}{dz} + \frac{\ii k_x^2}{D}\left(\ii c_s^2 \frac{d\ln\rho}{dz}
    - \frac{r}{k_x}\frac{d\Omega^2}{dz}\right)\delta v_z,\label{adia_iso1}\\
  & -\frac{\sigma^2}{c_s^2}\delta v_z = \gamma \frac{d\Delta}{dz} +
  \frac{d}{dz}\left(\frac{d\ln\rho}{dz}\delta v_z\right) +
  \left(\gamma-1\right)\frac{d\ln\rho}{dz} \Delta.\label{adia_iso2}
\end{align}
%{\bf note: incompressible mode, set delta to zero. only works for
% constant gravity.}

We make the low-frequency approximation and the replacement
$D\to \Omega_k^2$ in Eq. \ref{adia_iso1}---\ref{adia_iso2}. For the
analysis to be carried out, the latter is a more limiting
approximation than for isothermal perturbations because we make the
replacement before combining the pair of equations above.

Here, we also use the fact that in the
global disk, 
\begin{align}
  r\frac{\p \Omega^2}{\p z} = -\frac{\p \ln\rho}{\p z} \frac{qc_s^2}{r},
\end{align}
for the power-law prescription of the midplane sound-speed for
$\Gamma=1$.

Making the above substitutions and combining Eq. \ref{adia_iso1}---
Eq. \ref{adia_iso2} we obtain, in terms of dimensionless variables defined
previously, an equation for $\delta v_z$, 
\begin{align}
  0 =& \dd v_z^{\prime\prime} + \left(1 + \ii \epsilon q
    \hat{k}\right)\left(\ln\rho^{\prime}\delta v_z\right)^\prime\notag\\
  &+\left\{\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right)\right.\notag\\
  &\phantom{0=}\left.-\left(\frac{\gamma-1}{\gamma}\right)\left[\ln\rho^{\prime\prime}+\hat{k}^2\left(1-\frac{\ii\epsilon  
          q}{\hat{k}}\right)\ln\rho^{\prime 2}\right]\right\}\delta v_z.\label{adia_iso3}
\end{align}

We multiply Eq. \ref{adia_iso3} by $\rho\delta v_z^*$ and
integrate vertically, assume boundary terms vanish when integrating by
parts, to obtain
\begin{align}
  &\hat{\sigma}^2\left(\frac{1}{\gamma} +
    \hat{k}^2\right)\int_{\zhat_1}^{\zhat_2}\rho|\delta
  v_z|^2 d\zhat \notag\\
  &=  \left(\frac{\gamma-1}{\gamma}\right)
  \int_{\zhat_1}^{\zhat_2}\rho|\delta v_z^\prime|^2 d\zhat
  +\frac{1}{\gamma}\int_{\zhat_1}^{\zhat_2}\frac{1}{\rho}|(\rho\delta
  v_z)^\prime|^2 d\zhat\notag\\
  &\phantom{=} +
  \left(\frac{\gamma-1}{\gamma}\right)\hat{k}^2\left(1-\frac{\ii\epsilon
      q}{\hat{k}}\right) \int_{\zhat_1}^{\zhat_2}\rho\ln\rho^{\prime
    2}|\delta v_z|^2 d\zhat\notag\\
  &\phantom{=} + \ii\epsilon q \hat{k}
  \int_{\zhat_1}^{\zhat_2}\ln\rho^\prime(\rho\delta v_z^*)^\prime
  \delta v_z d\zhat.\label{adia_integral}
\end{align}
When $q\equiv0$, all the terms on the right-hand-side (RHS) are real. Then
$\hat{\sigma}^2$ is real, and  $\hat{\sigma}^2>0$ if $\gamma>1$. As
expected, a sub-adiabatically stratified disk is stable in the absense
of vertical shear. In the presence of vertical shear $q\neq0$ and
$\hat{\sigma}$ is generally complex, which implies the possibility of
instability.
